Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-10 csf-2017/paper.tex
suggestion for paper edition of f1 lemma
short text about f1 lemma proof

For both programs, we can prove that if they are linked up with an
adversary that is allowed to allocate new memory but otherwise has no
capabilities, then executing the program, the assertion will never
fail.
We can only prove this, because the capability machine encapsulates
the local state of the two programs. The two examples also illustrate
the versatility of the logical relation. The two examples use
different calling convention, but the logical relation is not specific
to any calling convention, so we can use it to reason about both
programs. In order to reason about the unknown untrusted part, we use
the Fundamental Theorem to argue that it must be capability safe.

\todo[inline]{give a more detailed high-level taste of a proof here? Maybe later for one of the other examples?}

\begin{lemma}[Correctness lemma for \texttt{f1}]
  \label{lem:correctness-f1}
  Let
  \begin{align*}
    c_{\var{adv}} & \defeq ((\entry,\glob),\start_{\adv},\addrend_{\adv},\start_{\adv}+\olf) \\
    c_{f1} & \defeq ((\rwx,\glob),\mathtt{f1}-\olf,\mathtt{1f},\mathtt{f1}) \\
    c_\malloc & \defeq ((\entry,\glob),\start_\malloc,\addrend_\malloc,\start_\malloc+\olf) \\
    c_{\var{stk}} & \defeq ((\rwlx,\local),\start_\stk,\addrend_\stk,\start_\stk-1) \\
    c_\link & \defeq ((\readonly,\glob),\link,\link+1,\link)\\
    \reg & \in \Regs \\
    m & \defeq \hs_{f1} \uplus 
        \hs_\flag \uplus                
        \ms_{\var{link}} \uplus 
        \hs_\adv \uplus 
        \ms_{\malloc} \uplus 
        \ms_{\var{stk}} \uplus
        \ms_{\var{frame}} 
  \end{align*}
  and
  \begin{itemize}
  \item $c_\malloc$ satisfies the specification for malloc and $\iota_{\malloc,0}$ is the region from the specification.
  \end{itemize}
  where 
  \begin{align*}
    &\dom(\hs_{f1}) = [\mathtt{f1}-\olf,\mathtt{1f}] \\
    &\dom(\hs_\flag) = [\flag,\flag] \\
    &\dom(\ms_\link) = [\link,\link+1]\\
    &\dom(\ms_\stk) = [\start_\stk, \addrend_\stk]\\
    &\dom(\hs_{\adv}) = [\start_\adv,\addrend_\adv] \\
    &\heapSat[\hs_{\malloc}]{n}{[0 \mapsto \iota_{\malloc,0}]} \qquad \text{ for all $n \in \nats$}
  \end{align*}
  and
  \begin{itemize}
  \item $\ms_{f1}(\mathtt{f1}-\olf) = ((\readonly,\glob),\link,\link+1,\link)$, $\ms_{f1}(\mathtt{f1}-\olf+1) = ((\readwrite,\glob),\flag,\flag,\flag)$, the rest of $\hs_{f1}$ contains the code of $f1$.
  \item $\ms_\flag = [\flag \mapsto 0]$
  \item $\ms_{\var{link}} = [\var{link} \mapsto c_\malloc, \var{link} + 1 \mapsto c_\adv]$
  \item $\hs_\adv(\start_\adv) = c_\link$ and $\forall \addr \in [\start_\adv+1,\addrend]\ldotp \ms_\adv(a) \in \ints$
  \end{itemize}
  if 
  \[
    (\reg\update{\pcreg}{c_{f1}}\update{r_\stk}{c_\stk},m) \step[n] (\halted,m'),
  \]
  then
  \[
    m'(\flag) = 0
  \]  
\end{lemma}

>>>>

For both programs, we can prove that if they are linked up with an
adversary, $\adv$, that is allowed to allocate new memory but otherwise has no
capabilities, then executing the program, the assertion will never
fail.
We can only prove this, because the capability machine encapsulates
the local state of the two programs. The two examples also illustrate
the versatility of the logical relation. The two examples use
different calling convention, but the logical relation is not specific
to any calling convention, so we can use it to reason about both
programs.

\todo[inline]{give a more detailed high-level taste of a proof here? Maybe later for one of the other examples?}
\begin{lemma}[Correctness lemma for \texttt{f1}]
\todo[inline]{The lemma copied directly from the TR. To be deleted later!}
  \label{lem:correctness-f1}
  Let
  \begin{align*}
    c_{\var{adv}} & \defeq ((\entry,\glob),\start_{\adv},\addrend_{\adv},\start_{\adv}+\olf) \\
    c_{f1} & \defeq ((\rwx,\glob),\mathtt{f1}-\olf,\mathtt{1f},\mathtt{f1}) \\
    c_\malloc & \defeq ((\entry,\glob),\start_\malloc,\addrend_\malloc,\start_\malloc+\olf) \\
    c_{\var{stk}} & \defeq ((\rwlx,\local),\start_\stk,\addrend_\stk,\start_\stk-1) \\
    c_\link & \defeq ((\readonly,\glob),\link,\link+1,\link)\\
    \reg & \in \Regs \\
    m & \defeq \hs_{f1} \uplus 
        \hs_\flag \uplus                
        \ms_{\var{link}} \uplus 
        \hs_\adv \uplus 
        \ms_{\malloc} \uplus 
        \ms_{\var{stk}} \uplus
        \ms_{\var{frame}} 
  \end{align*}
  and
  \begin{itemize}
  \item $c_\malloc$ satisfies the specification for malloc and $\iota_{\malloc,0}$ is the region from the specification.
  \end{itemize}
  where 
  \begin{align*}
    &\dom(\hs_{f1}) = [\mathtt{f1}-\olf,\mathtt{1f}] \\
    &\dom(\hs_\flag) = [\flag,\flag] \\
    &\dom(\ms_\link) = [\link,\link+1]\\
    &\dom(\ms_\stk) = [\start_\stk, \addrend_\stk]\\
    &\dom(\hs_{\adv}) = [\start_\adv,\addrend_\adv] \\
    &\heapSat[\hs_{\malloc}]{n}{[0 \mapsto \iota_{\malloc,0}]} \qquad \text{ for all $n \in \nats$}
  \end{align*}
  and
  \begin{itemize}
  \item $\ms_{f1}(\mathtt{f1}-\olf) = ((\readonly,\glob),\link,\link+1,\link)$, $\ms_{f1}(\mathtt{f1}-\olf+1) = ((\readwrite,\glob),\flag,\flag,\flag)$, the rest of $\hs_{f1}$ contains the code of $f1$.
  \item $\ms_\flag = [\flag \mapsto 0]$
  \item $\ms_{\var{link}} = [\var{link} \mapsto c_\malloc, \var{link} + 1 \mapsto c_\adv]$
  \item $\hs_\adv(\start_\adv) = c_\link$ and $\forall \addr \in [\start_\adv+1,\addrend]\ldotp \ms_\adv(a) \in \ints$
  \end{itemize}
  if 
  \[
    (\reg\update{\pcreg}{c_{f1}}\update{r_\stk}{c_\stk},m) \step[n] (\halted,m'),
  \]
  then
  \[
    m'(\flag) = 0
  \]  
\end{lemma}

In order to formulate a lemmas about \texttt{f1} and \texttt{f2}, we need a way to observe whether the assertion fails. To this end, we assum access to a flag which will be set by the assertion if it fails. A slightly simplified version of the correctness lemma for \texttt{f1} is:
\begin{lemma}
  \label{lem:correctness-f1}
  Let
  \begin{align*}
    c_{\var{adv}} & \defeq ((\entry,\glob),\dots) \\
    c_{f1} & \defeq ((\rwx,\glob),\dots) \\
    c_\malloc & \defeq ((\entry,\glob),\dots) \\
    c_{\var{stk}} & \defeq ((\rwlx,\local),\dots) \\
    c_\link & \defeq ((\readonly,\glob),\dots)\\
    \reg & \in \Regs \\
    m & \defeq
        \begin{aligned}[t]
          &\hs_{f1} \uplus 
        \hs_\flag \uplus                
        \ms_{\var{link}} \uplus 
        \hs_\adv \uplus \\
        &\ms_{\malloc} \uplus 
        \ms_{\var{stk}} \uplus
        \ms_{\var{frame}}
      \end{aligned}
  \end{align*}
  where each of the capability has an appropriate range of authority and pointer and
  \begin{itemize}
  \item $\ms_{f1}$ contains $c_\link$, $c_\flag$ and the code of \texttt{f1}
  \item $\ms_\flag(\flag) = 0$
  \item $\ms_{\var{link}}$ contains $c_\adv$ and $c_\malloc$
  \item $\hs_\adv$ contains $c_\link$ and otherwise only instructions.
  \end{itemize}
  if 
  \[
    (\reg\update{\pcreg}{c_{f1}}\update{r_\stk}{c_\stk},m) \step[*] (\halted,m'),
  \]
  then
  \[
    m'(\flag) = 0
  \]  
\end{lemma}

In order for prove Lemma~\ref{lem:correctness-f1}, it suffices to show
that the start configuration is capability safe (in the $\observations$ relation) for a
world with a permanent region that requires the assertion flag to be
0. This is done using an anti-reduction lemma which states that for a
configuration to be capability safe, it is sufficient to
show that it is in it at a later point of the execution. To reason
about \texttt{scall}, it suffices to show that the configuration just
after we jumped is capability safe and that the configuration just
after \texttt{scall} is done cleaning up is capability safe.
We use the Fundamental Theorem to reason about the unknown code, but
notice that the Fundamental Theorem says nothing about enter
capabilities and that the adversary capability is an enter
capability. Luckily the enter capability becomes a read/execute
capability when we jump and the Fundamental Theorem can be used to
show that this capability is capability safe as a program counter.
